Annotating, understanding, and predicting long-term video memorability

### 2018-02-01 ARTICLE.tex
Final version of article, submitted to ICMR

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{booktabs} % For formal tables
%% perso packages:
\usepackage{epstopdf}
\usepackage{microtype}
\usepackage{tabularx}
\usepackage{multirow}
%\usepackage{arydshln}
\usepackage{subfig}
\usepackage{balance}
\usepackage{caption}
\usepackage{graphicx}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CONFERENCE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Conference
\acmConference[ICMR 2018]{ACM Yokohama conference}{June 2018}{Yokohama, Japan} 
\acmYear{2018}
\copyrightyear{2018}

\acmArticle{4}%??
\acmPrice{15.00}%??

@@

%%%%%%%%%%%%
%%% for blind review %%%
%author 1
\author{Removed for review}
%author 2
\author{Removed for review}
%author 3
\author{Removed for review}
%author 4
\author{Removed for review}

% The default list of authors is too long for headers.
\renewcommand{\shortauthors}{R. Cohendet et al.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
The growing number of video contents on the Internet encourages us to find new ways to make more relevant their occurrences in our everyday life.
In this context, memorability can be regarded as a useful metric of video importance to help us make a choice between competing videos.
However, the research on computational understanding of video memorability is in its early stages.
There is no available dataset for modelling purposes, and the few previous attempts provided protocols not generalizable to collect data at a large scale.
Furthermore, the video computational features valuable to build a robust memorability predictor remain largely undiscovered.
In this article, we propose a new protocol to collect long-term memorability annotations, that we use to measure memory performances of 104 participants from weeks to years after memorization.
We then analyze the data collected for 660 videos by focusing on its quality, and test it against different characteristics such as response time, duration of memory retention and repetition of visualization.
We finally conduct an extensive feature analysis, comparing several methods and classes of features, to propose a computational model for the prediction of video memorability.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% KEYWORDS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\keywords{Video memorability, Long-term memory, Measurement protocol, Memorability scores, Deep learning, Multimedia information retrieval}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}%nb of sections %2 PAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%opening
Enhancing the relevance of multimedia occurrences in our everyday life requires to imagine new ways to organize -- in particular, to retrieve -- digital contents.
Like other metrics of video "importance", such as aesthetics or interestingness, memorability can be regarded as useful to help us make a choice between competing videos.
In addition, memorability has the advantage of being clearly definable and objectively measurable (i.e., using a measure that is not influenced by the observer's personal judgment).
Thus, memorability has initially be defined as the probability for an image to be recognized \textbf{a few minutes} after a single view, when presented amidst a stream
of images \cite{isola_2011_makes}.
This definition has been widely accepted within subsequent work (e.g., \cite{mancas_2013_memorability,kim_2013_relative,celikkale_2013_visual,khosla_2015_understanding,lahrache_2016_bag}).

%from images to VM
The computational understanding of video memorability (VM) follows on from the study of image memorability prediction which has attracted increasing attention from the 2001 seminal work of Isola \textit{et al.} \cite{isola_2011_makes}.
With the recent introduction of deep learning to address the challenge of image memorability prediction, models also achieved very good results \cite{khosla_2015_understanding,baveye_2016_deep,squalli_2017_deep}.
As a result of this success, researchers have recently extended this challenge to the videos.
However, to the best of our knowledge, only two available studies focused on VM prediction \cite{han_2015_learning,shekhar_2017_show}.

%explain the scarcity of studies on VM
Several problems could explained this scarcity of studies.
Firstly, there is no publicly available dataset to train and test models.
This is probably the most serious obstacle to the rapid expansion of the VM prediction's search field.
Accordingly, to provide researchers with ground truth data should be our very first objective, similarly with the work of \cite{isola_2011_makes} and \cite{khosla_2015_understanding} which have enabled research on image memorability to flourish.
The second point is closely related to the first one: there is no widely accepted definition of VM.
The previous attempts to predict VM \cite{han_2015_learning,shekhar_2017_show} were based on different measures of memorability.
Furthermore, contrary to images, videos do not constitute clearly defined units.
They have supplementary dimensions -- sound and movement -- that critically contribute to the semantic and emotional information conveyed.
If harmonized, the videos used and the way memorability is measured will have a critical impact on what we will understand by VM.
The definition of image memorability by \cite{isola_2011_makes} had a great impact on subsequent work as aforementioned.
But it also inevitably limited researchers.
In particular, in previous research image memorability corresponds to memory performances measured only a few minutes after memorization.
But these might be poor predictors of longer term memory performances (at least in some instances; e.g., for emotional images \cite{cohendet_2016_prediction}).
Thus, VM data would benefit from a protocol that would measure lasting long-term memory performance.

%modelling
Regarding modelling, our capacity to propose efficient computational models is also important to meet the challenge of VM prediction. 
The previous attempts at predicting VM \cite{han_2015_learning,shekhar_2017_show} shed light on several features which have a predictive power of VM.
However, videos contain a great volume of data which can be used for feature extraction purposes, and the work is far for complete.

%objectives of the present study
The aim of this article is to participate in the expansion of the VM prediction emerging search field.
In particular, we:
\begin{itemize}
	\item propose a new protocol that measures very long-term memory performances (from weeks to years after memorization) to collect quality ground truth data (section 2);
	\item assess and analyze the collected data to better understand VM (section 3);
	\item model VM in an attempt to understand which computation features are better for this task (section 4).
\end{itemize}
In the next subsection, we review previous work on image and video memorability prediction, focusing on protocol and modelling questions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Previous work}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Measurement of memorability}
%Images work -- towards a long-term memorability
Almost all studies on image memorability prediction made use of one or the other of the two available large datasets designed specifically to meet this challenge \cite{isola_2011_makes,khosla_2015_understanding}.
To build these datasets, the authors, possibly constrained by the difficulty in conducting long crowdsourcing studies, measured memory performance a few minutes after memorization to obtain their memorability annotations.
As proposed in \cite{cohendet_2016_prediction}, this could be a problem if we conceive that memorability reflects a lasting memory performance.
Indeed, it has long been shown that long-term memories continue to change long after their memorization through an ongoing process called consolidation, which lasts weeks to years \cite{mcgaugh_2000_memory}.
%In particular, the early work of the psychologist Ebbinghaus showed that the drop in long-term memory performance in recall follow an exponential decay, and is particularly strong in the few days immediately following the memorization \cite{ebbinghaus_1913_memory}.
Because several factors influence the consolidation process (e.g., emotions, sleep, re-evocation), which does not equally affect all memories, the order of memorability ratings measured for videos is susceptible to change over time \cite{cohendet_2016_prediction}.
A protocol to collect memorability annotations would benefit from the capacity to capture long-lasting memory performances, averaged to obtain what we will later refer to as "long-term memorability".

%Han et al.
To our knowledge, the first attempt at predicting VM had been proposed by \cite{han_2015_learning}. 
The authors partially adapted the protocol proposed by \cite{isola_2011_makes} to measure image memorability for videos.
In contrast to the "memory game" proposed by Isola \textit{et al.} to collect memorability data, their protocol is however much heavier.
They used a classical recognition task to measure memory for videos, which consists in two steps: a free viewing task, followed two days later by a recall task.
The task duration, for each of the 20 participants, was about 24 hours, spread over 10 sessions (five free viewing tasks and five recall tasks) of about two hours each.
The authors used the same proportion of fillers (i.e., non repeated videos) in the free viewing and recall tasks (i.e., 4/5 of fillers and 1/5 of repeated videos named targets) to, according to them, guarantee that viewers were unaware of targets.
If it was mandatory in \cite{isola_2011_makes} for which encoding and recall tasks were interlaced, there is here a way to alleviate the task without impacting its quality; indeed, reducing the number of fillers in the free viewing task would have very little impact on the memorability scores (often authors even use only material interrogated later in the learning/free viewing task).
Furthermore, as pointed by \cite{shekhar_2017_show}, the long time span of the experiment makes difficult the generalization of this protocol, in particular to build an extensive dataset.
Furthermore, authors measure memory after two days, but the consolidation process lasts weeks to years: it would be interesting to collect even longer term memory performance measures.

%Shekhar et al.
Another earlier approach was the one of \cite{shekhar_2017_show}.
The participants performed a crowdsourcing experiment consisted of a free viewing task where they saw a sequence a videos, followed by a recall task for which they had to answer textual question (such as: "Did you see a man juggling?", "Did you see a car on a road?").
The major drawback of the study comes from the use of questions instead of a classic visual recognition task.
Indeed, the memorability scores computed for the videos may reflect not only the differences in memory performances but also the differences between the questions in terms of difficulty
The authors have tested complexity of textual questions, but their measure is not sufficient to represent the real difficulty of the question in its complexity (ease of understanding, in particular for the non-anglophone people that works with Amazon Mechanical Turk, ease of imaging, ease to retrieve a scene by some words more the by others...).
Especially since the authors took into account the response time of the participants to calculate the memorability scores of the videos.
One will note that this potential bias could have also affected the measure of inter-human consistency.
Furthermore, the questions are handcrafted, and the choices for types of questions and videos are very limited (e.g., the question "Did you see a car on a road?" implies that in a whole experimental session one can have just one car on a road).
The authors also manually ensured that no textual questions nor videos in a session were similar in content.
This makes it difficult to generalize this protocol to the construction of an extensive dataset.

\subsection{Memorability modelling}
%intro on focus on features to discover
Previous attempts at predicting image and video memorability highlighted quite a few features that correlates with memorability.

%image memorability prediction
The pioneering work of Isola \textit{et al.} focused primarily on building computational models to predict image memorability from low-level visual features \cite{isola_2011_makes}.
It appeared from this first attempt that it is possible to predict to some extent the degree of picture's memorability.
Several characteristics have been found to be relevant for predicting memorability in subsequent work, for example saliency features \cite{mancas_2013_memorability}, interestingness and aesthetic \cite{isola_2014_makes}, or emotions \cite{khosla_2015_understanding}.
The best results was finally obtained by using fine-tuned deep features, which outperformed all other features in \cite{khosla_2015_understanding}, reaching a rank correlation of $.64$ which is near human consistency ($.68$) measured for ground truth collected in the study.
This result was confirmed in \cite{baveye_2016_deep,squalli_2017_deep}.

%VM prediction
Regarding the VM prediction, Han \textit{et al.} propose a method which combines the power of audiovisual and fMRI-derived features \cite{shekhar_2017_show}.
They preliminary built a computational model learned with fMRI features, which supposedly convey the brain activity of memorizing videos, which enable them to finally predict VM without fMRI scans.
However, the method would be difficult to generalize

Shekhar \textit{et al.} conducted a performance analysis of several computationally extracted features before building their memorability predictor \cite{shekhar_2017_show}.
The analysis encompassed C3D deep learning feature for video classification, video semantics obtained thanks to video captioning method, saliency features, dense trajectories, and color features.
They found that the most predictive feature combination used video semantics, spatio-temporal, saliency and color features.
The feature that performed the best when tested alone was video semantics.
The authors used a captioning method to generate a semantic description of the videos.
The text generated by this method was then processed by a recursive auto-encoder network that outputted a 100-dimensional representation of the videos.
Due to the particularity of the dataset of Shekhar \textit{et al.} -- that is, their aforementioned use of questions to measure memorability --, it would be interesting to confirm if this combination works equally well on another dataset, and in particular if images captioning features also perform the best.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Memorability dataset construction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Video collection}
%why did we choose movie
We wanted our protocol to measure memory performance after a significant retention period.
This can be achieved either with a longitudinal study, or by measuring a memory created prior to the experiment.
We chose the latter because it enabled us to immediately measure very long-term memory.
Thus, the main characteristic of the proposed protocol, in contrast with previous work, is the absence of learning (often, free viewing) task, replaced by a questionnaire designed to collect information about the participants' prior memory.

%movies and videos selections
 %with HD or dvdrip quality, in orginal (i.e.,english) version, without without subtitles [add if results HD vs. DVDrip]
We first established a list of 100 movies, taking care to mix popularity and genres.
We then manually selected seven videos of 10 seconds from each movie.
To maintain a high intra-video semantic cohesion, we did not make cuts that would impair the understanding of the scene, nor did we aggregate shots that belong to different scenes.
Indeed, since the semantics is linked to the memorability of images \cite{isola_2014_makes}, we can expect it is linked to the memorability of videos too.

%neutral vs. typical
We also gave preference to the videos we called "neutral", by contrast to the "typical" ones.
According to our definition, a neutral video is a part of a movie which contains no element that would enable someone to easily guess it belongs to a particular movie.
The list of undesirable elements includes but is not limited to: recognizable famous actors, typical music, style, etc.
Typical videos are simply defined as non-neutral videos.
In most movies, just a few or no 10-sec neutral videos exist.
That explains why we obtained only 127 neutral videos for 573 typical ones (while we expected two neutral and five typical videos per movie).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PACKAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{booktabs} % For formal tables
%% perso packages:
\usepackage{epstopdf}
\usepackage{microtype}
\usepackage{tabularx}
\usepackage{multirow}
%\usepackage{arydshln}
\usepackage{subfig}
\usepackage{balance}
\usepackage{caption}
\usepackage{graphicx}
\usepackage{amsmath}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CONFERENCE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Conference
\acmConference[ICMR 2018]{ACM Yokohama conference}{June 2018}{Yokohama, Japan} 
\acmYear{2018}
%\copyrightyear{2018}
%\acmArticle{4}%??
%\acmPrice{15.00}%??

@@

%%%%%%%%%%%%
%%% for blind review %%%
%author 1
%\author{Removed for review}
%author 2
%\author{Removed for review}

% The default list of authors is too long for headers.
\renewcommand{\shortauthors}{Removed for blind review}%{R. Cohendet et al.}%Removed for blind review

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
%The growing number of video contents on the Internet encourages us to find new ways to make more relevant their occurrences in our everyday life.
Memorability can be regarded as a useful metric of video importance to help make a choice between competing videos.
Research on computational understanding of video memorability is however in its early stages.
There is no available dataset for modelling purposes, and the few previous attempts provided protocols to collect video memorability data that would be difficult to generalize.
Furthermore, the computational features needed to build a robust memorability predictor remain largely undiscovered.
In this article, we propose a new protocol to collect long-term video memorability annotations.
We measure the memory performances of 104 participants from weeks to years after memorization to build a dataset of 660 videos for video memorability prediction.
This dataset is made available for the research community.
We then analyze the collected data in order to better understand video memorability, in particular the effects of response time, duration of memory retention and repetition of visualization on video memorability.
We finally investigate the use of various types of audio and visual features and build a computational model for video memorability prediction. We conclude that high level visual semantics help better predict the memorability of videos. 
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% KEYWORDS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\keywords{Video memorability, Long-term memory, Measurement protocol, Deep learning, Multimedia information retrieval}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}%nb of sections %2 PAGES
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%opening
Enhancing the relevance of multimedia occurrences in our everyday life requires new ways to organize -- in particular, to retrieve -- digital content.
Like other metrics of video "importance", such as aesthetics \cite{dhar_2011_high} or interestingness \cite{demarty_2016_mediaeval}, memorability can be regarded as useful to help make a choice between competing videos.
In addition, memorability has the advantage of being clearly definable and objectively measurable (i.e., using a measure that is not influenced by the observer's personal judgment).
Image memorability has initially been defined as the probability for an image to be recognized \textbf{a few minutes} after a single view, when presented amidst a stream
of images \cite{isola_2011_makes}.
This definition has been widely accepted within subsequent work (e.g., \cite{mancas_2013_memorability,kim_2013_relative,celikkale_2013_visual,khosla_2015_understanding,lahrache_2016_bag}).

%from images to VM
The computational understanding of video memorability (VM) follows on from the study of image memorability (IM) prediction which has attracted increasing attention since the seminal work of Isola \textit{et al.} \cite{isola_2011_makes}.
With the recent introduction of deep learning to address the challenge of IM prediction, models also achieved very good results \cite{khosla_2015_understanding,baveye_2016_deep,squalli_2017_deep}.
As a result of this success, researchers have recently extended this challenge to videos.
However, to the best of our knowledge, only two available studies focus on VM prediction \cite{han_2015_learning,shekhar_2017_show}.

%explain the scarcity of studies on VM
Several problems could explain this scarcity of studies on VM.
Firstly, there is no publicly available dataset to train and test models.
This is probably the most serious obstacle to the rapid expansion of research in VM prediction.
Following the foot steps of researchers in IM \cite{isola_2011_makes,khosla_2015_understanding}, providing data and ground truth for VM should be one of our first objectives.
The second point, closely related to the previous one, is the lack of a common definition for VM.
The previous attempts to predict VM \cite{han_2015_learning,shekhar_2017_show} were based on different measures of memorability.
Furthermore, in comparison to images, videos have supplementary dimensions -- sound and visual movement -- that critically contribute to the semantic and emotional information conveyed and this makes it difficult to come up with a common definition of VM.
But the videos used and the way memorability is measured have a critical impact on what we understand by VM.
Similarly, the definition of IM \cite{isola_2011_makes} inevitably limited researchers.
In particular, previous research on IM focused on the measurement of memory performances only a few minutes after memorization.
But these short-term memory performances might be poor predictors of longer term memory performances.
For example, the memorability of emotional images would decrease in a non-harmonious way between two measures of memory performance (a few minutes and one day after memorization), inducing in a dataset a change in the ranking of memorability values over time \cite{cohendet_2016_prediction}.
To this end, VM data would be expected to benefit from a protocol that would measure lasting long-term memory performance.

%modelling
Regarding modelling, the previous attempts at predicting VM \cite{han_2015_learning,shekhar_2017_show} highlighted several features which contribute to the prediction of VM, such as semantic, saliency and color features.
But the work is far from complete and our capacity to propose efficient computational models will participate to answer the challenge of VM prediction.

%objectives of the present study
While participating in the expansion of research on VM, the contributions of this paper are threefold:
\begin{itemize}
	\item We propose a new protocol that measures very long-term memory performances (from weeks to years after memorization) to collect ground truth data of good quality and make the corresponding dataset available for the research community (section \ref{section_dataset_construction}).
	\item We assess and analyze the collected data and provide useful insights on the understanding of VM (section \ref{study_memorability}).
	\item We build a computational model based on machine learning techniques which allows to predict VM score of a given video. For this purpose, we investigate the use of various types of audio and visual features, ranging from low-level characteristics to emotional and (highly) semantic features (section \ref{mem-pred}).
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Previous work}
In this section, we review previous work on image and video memorability prediction, focusing on annotation protocols and modelling aspects.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Measurement of memorability}
%work in psychology
Long-term memory has been studied for over a century in psychology, from multiple perspectives, since the seminal experimental studies of Ebbinghaus \cite{ebbinghaus_1913_memory} until the more recent neuro-imaging studies \cite{blumenfeld_2007_prefrontal}.
This work provided researchers, interested in computational understanding of IM and VM, with several memory tests (see \cite{richardson_1988_measures} for an extensive overview), such as the recognition test \cite{isola_2011_makes,khosla_2015_understanding,han_2015_learning} or the textual questions-based recall survey \cite{shekhar_2017_show}.
It also demonstrated that humans have an extensive long-term visual memory, which enables them to recall a great amount of images \cite{standing_1973_learning} and image details \cite{brady_2008_visual}, as well as of videos \cite{furman_2007_they}.
This is one of the observations at the origin of the work on memorability in computer science \cite{isola_2011_makes}.
Several factors have also been highlighted for their critical influence on long-term memory, such as emotion \cite{kensinger_2008_memory}, semantics \cite{quillan_1966_semantic}, several demographic factors \cite{cohendet_2016_using}, memory re-evocation \cite{nadel_1997_memory}, or passage of time \cite{mcgaugh_2000_memory}.
These different factors are all important to better understand memorability, and are found valuable computational features for image and video memorability prediction.

%Images work -- towards a long-term memorability
Focusing on the work in computer vision, most studies on IM prediction made use of one of the two available large datasets designed specifically to meet this challenge \cite{isola_2011_makes,khosla_2015_understanding}.
To address more specific problems, several other datasets have also been publicly released concerning memorability of face photographs \cite{bainbridge_2013_intrinsic}, visualization pictures \cite{borkin_2013_makes}, emotional images \cite{cohendet_2016_using} and scene categories \cite{bylinskii_2015_intrinsic}.
To build the most used datasets presented in \cite{isola_2011_makes,khosla_2015_understanding}, the authors, possibly constrained by the difficulty in conducting long crowdsourcing studies, measured memory performance a few minutes after memorization to obtain their memorability annotations.
This could be a problem if we conceive that memorability reflects a lasting memory performance \cite{cohendet_2016_prediction}.
Indeed, it has been shown that long-term memories continue to change long after their memorization through an ongoing process called consolidation, which lasts weeks to years \cite{mcgaugh_2000_memory}.
%In particular, the early work of the psychologist Ebbinghaus showed that the drop in long-term memory performance in recall follow an exponential decay, and is particularly strong in the few days immediately following the memorization \cite{ebbinghaus_1913_memory}.
Because several factors influence the consolidation process (e.g., emotions, sleep, re-evocation), which does not equally affect all memories, the order of memorability ratings measured for content, and especially videos, is susceptible to change over time \cite{cohendet_2016_prediction}.
A protocol to collect memorability annotations would benefit from the capacity to capture long-lasting memory performances, averaged to obtain what we will later refer to as "long-term memorability".

%Han et al.
To our knowledge, the first attempt at measuring VM \cite{han_2015_learning} partially adapted the protocol proposed to measure IM \cite{isola_2011_makes} to videos.
The resulting protocol is however much heavier than the memory game protocol \cite{isola_2011_makes}.
They followed a classical recognition process, which consists of two steps: a free viewing task, followed two days later by a recall task.
The task duration, for each of the 20 participants, was about 24 hours, spread over 10 sessions (five free viewing tasks and five recall tasks) of about two hours each.
The authors used the same proportion of fillers (i.e., non repeated videos) in the free viewing and recall tasks (i.e., 4/5 of fillers and 1/5 of repeated videos named targets) to guarantee that viewers were unaware of targets.
If it was mandatory in \cite{isola_2011_makes} for which encoding and recall tasks were interlaced, there is a way here to alleviate the task without impacting its quality; indeed, reducing the number of fillers in the free viewing task would have very little impact on the memorability scores (often authors even use only material interrogated later in the learning/free viewing task).
Furthermore, the long time span of the experiment makes the generalization of this protocol difficult, in particular if one targets the construction of an extensive dataset.
Moreover, authors measured memory after two days, but, as aforementioned, it is known that the consolidation process lasts weeks to years: it would be interesting to collect even longer term memory performances.

%Shekhar et al.
In another earlier approach for VM measurement, the participants performed a crowdsourcing experiment that consisted of a free viewing task during which they saw a series of videos, followed by a recall task in which they had to answer textual questions (such as: "Did you see a man juggling?", "Did you see a car on road?") \cite{shekhar_2017_show}.
The major drawback of the study comes from the use of questions instead of a classic visual recognition task.
Indeed, the memorability scores computed for the videos may reflect not only the differences in memory performances but also the differences between the questions in terms of difficulty.
The authors tested the complexity of the chosen textual questions using the Flesch-Kincaid Grade Level readability metric, which is designed to quantify how difficult to understand an English text is \cite{kincaid_1975_derivation}.
However, this measure might not be sufficient to represent the questions in all their complexities, such as the ease for a person to draw images from the words, to connect the words to the associated concepts in memory, to associate the words to the corresponding scene, etc.
Moreover, there is an unequal distribution of the reading comprehension of English among Amazon Mechanical Turk workers, who are not necessary anglophone people.
This problem of a potential unequal complexity of the question becomes even more important in view of the use by authors of the response time to calculate memorability scores, which might also critically depend on the complexity of the questions.
One will note that this potential bias could have also affected the measure of inter-human consistency.
Furthermore, the questions were handcrafted, and the choices for types of questions and videos were very limited (e.g., the question "Did you see a car on road?" implies that in a whole experimental session only one car on a road should appear to connect the memory performance to one particular video).
The authors also manually ensured that no textual questions nor videos in a session were similar in content.
This makes it difficult to generalize this protocol to the construction of an extensive dataset.

\subsection{Memorability modelling}
%intro on focus on features to discover
Previous attempts at predicting image and video memorability highlighted quite a few features that correlate with memorability.

%IM prediction
The pioneering work of Isola \textit{et al.} focused primarily on building computational models to predict IM from low-level visual features \cite{isola_2011_makes}.
From their work, it appeared that the degree of an image's memorability can be predicted to a certain extent.
Several characteristics have also been found to be relevant for predicting memorability in subsequent work, for example saliency \cite{mancas_2013_memorability}, interestingness and aesthetics \cite{isola_2014_makes}, or emotions \cite{khosla_2015_understanding}.
The best results were finally obtained by using fine-tuned deep features, which outperformed all other features in \cite{khosla_2015_understanding}, reaching a rank correlation of $.64$ which is near human consistency ($.68$) when measured for the ground truth collected in the study.
This result was later confirmed by other researchers \cite{baveye_2016_deep,squalli_2017_deep}.

%VM prediction
Regarding VM prediction, Han \textit{et al.} proposed a method which combines the power of audio-visual and fMRI-derived features \cite{han_2015_learning}.
They preliminary built a computational model learned with fMRI features, which supposedly conveys the brain activity of memorizing videos.
This enabled them to finally predict VM without the use of fMRI scans in a second step.
However, the method would be difficult to generalize.
Shekhar \textit{et al.} conducted a performance analysis of several computationally extracted features before building their memorability predictor \cite{shekhar_2017_show}.
The analysis encompassed C3D deep learning features, semantic features obtained from some video captioning process, saliency features, dense trajectories, and color features.
They found that the most predictive feature combination used captioning features, dense trajectories, saliency and color features.
The features that performed the best when used alone were image captioning features, i.e., those conveying more semantics.
Due to the particularity of the dataset of Shekhar \textit{et al.} -- that is, their aforementioned use of questions to measure memorability --, it would be interesting to confirm if this combination works equally well on another dataset, and in particular if image captioning features also perform the best.
